Fast Sorting on Modern Computers

### 2005-08-16 docs/tech_report/merge.tex
removed the word project, and resolved the occasional TODO
merged davids final changes in

While this step makes sense, whether this is valuable is very much a platform
dependant issue. Since virtual addresses are used, it may be possible that the
addresses used by the cache are not aligned as expected. Even worse, an
operating system with an excellent memory management system may do a similar
step itself. The program's attempt to reduce cache misses could then increase
them. In the event that the \n{Memory Management Unit} in the processor uses
virtual addresses to index its cache, or has a direct mapping between virtual
and physical or bus addresses, then this problem is very unlikely. The
simulations presented in this project address the cache by virtual address, and
so this optimization is effective in this case.

@@

Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. The
labels in the graphs correspond to the names of the steps in Knuth's algorithm,
which also correspond with the labels in our code.

Base mergesort has the same number of mispredictions as Knuth's sorts (see
Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
`\n{Right}' refer to iterating through the first (or left-hand) and second (or
right-hand) lists being compared, respectively.

Base mergesort has a lower number of branches than Knuth's sorts, probably due
to performing a presort first. The presort has approximately 6.8 misses per key
(not shown in the graphs). In later mergesorts, an insertion sort is used
instead.

Double-aligned tiled mergesort (which has the same results are tiled mergesort)
has a predilection for merging forwards, shown in \ref{mergesorts-4}. The
`\n{Left}' and `\n{Right}' in the labels refer to the same as in
\ref{mergesorts-3}, while `\n{Forwards}' and `\n{Reverse}' refer to whether the
first or second part of the bitonic array segment is being merged.

This predilection is likely to be due to the fact that the final merge goes
forwards. The left merge step is also favoured, but this is merely due to the
fact that it occurs first in the code. While there are more left branches, the
actual number of branches taken are the same.

In both double-aligned tiled mergesort and base mergesort, approximately two
thirds are of the loops are taken, unlike Knuth's sorts. However, this appears
to indicate an increase in branches, as opposed to a decrease in mispredictions.
Overall this stay is line with the idea that there is a minimum number of branch
misses, below which comparative sorts will not drop.

Double multi-mergesort sorts the arrays fully in the cache, leaving them
bitonically arranged. This means that there is no final merge, and hence no bias
for the forward branches, as there was in tiled mergesort. This is shown in
Figures \ref{mergesorts-5} and \ref{mergesorts-6}. The labels in these graphs
are the same as in Figure \ref{mergesorts-4} and \ref{base and memory-tuned
heapsorts-2}.

While merging, the number of branches is 20\% lower than base mergesort, while
the number of mispredictions in only 10\% lower. This implies that most of the
mispredictions occur early in the merging process, and not in the final merge,
which multi-mergesort leaves out.

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
heapsort curve}). Note that `\n{comparison 0}' occurs every time, and would be
optimized out by the compiler.

>>>>

While this step makes sense, whether this is valuable is very much a platform
dependant issue. Since virtual addresses are used, it may be possible that the
addresses used by the cache are not aligned as expected. Even worse, an
operating system with an excellent memory management system may do a similar
step itself. The program's attempt to reduce cache misses could then increase
them. In the event that the \n{Memory Management Unit} in the processor uses
virtual addresses to index its cache, or has a direct mapping between virtual
and physical or bus addresses, then this problem is very unlikely. The
simulations presented in this report address the cache by virtual address, and
so this optimization is effective in this case.

@@

Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
algorithm N.  The labels in the graphs correspond to the names of the steps in
Knuth's algorithm (we also used the same names for the labels of branch targets
in our C source code. Branch N3 is the main comparison branch; it is used to
choose the smaller of the elements from the start of the two sub-arrays being
merged. The direction of this branch is completely random -- it depends entirely
on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
this branch is almost entirely unpredictable using a bimodal predictor. The
result for a two-level predictor is much the same, because there is no pattern
in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
increasing elements) are control-flow branches, and are almost entirely
predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
Algorithm S is just as unpredictable as that in algorithm N. However, given that
the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
is possible to eliminate one of the control-flow branches.

Base mergesort is an optimized algorithm that reverses the sorting direction of
every second sub-array, pre-sorts small sub-arrays, and applies a number of
other optimizations. As Figure \ref{mergesorts-3} shows, our base mergesort has
eliminated some of the control-flow branches. The main comparison branch is
split into two separate branches --- one iterating from left-to-right over the
increasing sub-array (`\n{Left}'), and the other from right-to-left over the
decreasing sub-array (`\n{Right}').  However, base mergesort appears to need a
total of 50\% more comparison branches that either Algorithm S or N.

In fact, base mergesort eliminated control-flow branches by using the same
branch that is used for comparison to also deal with the case where one (or
both) of the merging sub-arrays becomes empty. Thus, the sum of the number of
executions of the branches `\n{Left}' and `\n{Right}' is almost exactly the same
as the sum of the number of executions of the branches N3 and N3 i==j in
Algorithm N. The total number of mispredictions is also almost the same.

The comparison branch in double-aligned tiled mergesort is divided into four
separate branches. As in base mergesort, there are `\n{Left}' and `\n{Right}'
branches. However, for each of these there are separate branches for when
creating a new sub-array in increasing order (`\n{Forwards}') or decreasing
order (`\n{Reverse}'). There are slightly more executions of `\n{Forwards}' than
`\n{Reverse}' because the final array is always sorted into increasing order.
The left merge step is also favoured, but this is merely due to the fact that it
occurs first in the code. While there are more left branches, the actual number
of branches taken are the same.

Double aligned multi-mergesort is an even more complicated algorithm.  It sorts
the arrays fully in the cache, leaving them bitonically arranged (sorted into
sub-arrays, sorted into alternatey increasing and decreasing order). The final
merge is performed separately, and hence no bias for the forward branches, as
there was in tiled mergesort. The final sort is performed with a $k$-way merge,
which is implemented using a heap data structure. This is shown in Figures
\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
comparisons from the merge. The labels in these graphs are the same as in Figure
\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.

During regular two-way merging, the number of branches is 20\% lower than base
mergesort, while the number of mispredictions in only 10\% lower. This implies
that most of the mispredictions occur early in the merging process, and not in
the final merge, which multi-mergesort leaves out.

Overall, when the $k$-ary heap is used, there are approximately the same total
number of branch mispredictions as in the other mergesort algorithms. Although
the final $k$-way merge executes a lot more comparison branches, these heap
branches are quite predictables, as we saw in Section
\ref{heapsort-branch-prediction}. This means that using the technique to reduce
cache misses does not adversely affect branch prediction rates (it does,
however, still affect the instruction count).  The rate of branch misses is much
the same as that in heapsort, and progresses in the same manner (see Figure
\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
illustration purposes only. It is guaranteed to be taken every time, and would
be optimized out by the compiler.
